\documentclass[graybox,sectrefs]{svmono_mod}
%\documentclass[graybox,envcountchap,sectrefs]{svmono_mod}
%\documentclass[envcountsame,envcountchap]{svmono}

% choose options for [] as required from the list in the Reference Guide
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{mathptmx}
\usepackage{helvet}
\usepackage{courier}
\usepackage{type1cm}         
\usepackage{makeidx}         % allows index generation
\usepackage{graphicx}        % standard LaTeX graphics tool
                             % when including figure files
\graphicspath{{Figures/}}
\usepackage{multicol}        % used for the two-column index
\usepackage[bottom]{footmisc}% places footnotes at page bottom
\usepackage{array,booktabs,calc}

\usepackage{listings}
\usepackage[final]{pdfpages}
% see the list of further useful packages in the Reference Guide

\usepackage{enumerate}
\usepackage{enumitem}

\usepackage{tikz,pgfplots}
\usetikzlibrary{arrows.meta,quotes}
\usepackage[mode=buildnew]{standalone}

\usepackage{bbold}
\usepackage{bm}
\usepackage{hyperref}
\usepackage{capt-of}

% Some modifications
\input{format.tex}

% Macros for easier typing
\input{main_definitions} 

\makeindex             % used for the subject index
                       % please use the style svind.ist with
                       % your makeindex program


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

\author{compiled by Jes\'us Cid-Sueiro}
\title{Stochastic Processes}
\subtitle{Year 2021-22}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Stochastic Processes}
\footnote{\textbf{Note}: This document has been compiled and adapted by Jesús Cid-Sueiro. Secs 1.1-1.4 are an adaptation of material taken from \cite{MITcourse} (Creative Commons license BY-NC-SA 4.0). Sec. 1.5 is a translation, summary and adaptation of a chapter in \cite{ComDig2007} made by the author with permission of the co-authors. Secs. 1.1-1.4 is published under license BY-NC-SA 4.0. Sec. 1.5 cannot be published or redistributed in any form. It cannot be transformed or used partially without permission of the original authors}
\vspace{11cm}

\mainmatter 
\tableofcontents

\include{Contents/stochastic_processes/MC0_IntroSP}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{thebibliography}{99.}

\bibitem{MITcourse}
Kempthorne, Peter, Choongbum Lee, Vasily Strela, and Jake Xia. "Topics in mathematics with applications in finance." Massachusetts Institute of Technology: MIT OpenCouseWare), https://ocw.mit.edu/courses/18-s096-topics-in-mathematics-with-applications-in-finance-fall-2013/ (Accessed Feb, 2023). License: Creative Commons BY-NC-SA.

\bibitem{ComDig2007}
Artés-Rodríguez, A., Pérez-González, F., Cid-Sueiro, J., López-Valcarce, R., Mosquera-Nartallo, C., Pérez-Cruz, F. (2012). Comunicaciones digitales (\url{http://www.tsc.uc3m.es/~antonio/libro_comunicaciones}).

\end{thebibliography}


\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




